---
title: 'One-Way ANOVA: Comparing Group Means'
author: "<br><br><span style='font-size:22px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong><br>College of Graduate Studies<br>Northeast Ohio Medical University</span>"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    transition: fade
    incremental: false
    code-fold: false
    code-line-numbers: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

```{=html}
<style>
.reveal h1, .reveal h2 { color: #2C7BE5; }
.reveal p, .reveal li { font-size: 0.86em; }
.small { font-size: 0.72em; }
.script { font-size: 0.74em; border-left: 5px solid #2C7BE5; padding-left: 0.8em; }
.callout { border-left: 6px solid #2C7BE5; padding-left: 0.8em; }
</style>
```

```{r setup, include=FALSE}
library(tidyverse)
library(broom)

theme_set(
  theme_minimal(base_size = 16) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.subtitle = element_text(size = 12),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 20, hjust = 1)
    )
)

set.seed(2026)
```

### Learning Objectives

-   define one-way ANOVA
-   describe outcome variable, factor, and factor levels within the context of one-way ANOVA.
-   explain the null and alternative hypotheses for a one-way ANOVA
-   interpret the components of the ANOVA table
-   fit one-way ANOVA model and evaluate the assumptions of the model

------------------------------------------------------------------------

### One-Way ANOVA : Definition

<br> <br>

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
<p align="center">

<strong>One-Way ANOVA</strong>

</p>

-   statistical method for comparing the means of a continuous outcome across groups defined by **one categorical factor**
-   evaluates whether the observed differences among sample means are large relative to the variability observed within groups
-   generalizes the two-sample mean comparison to settings with **three or more groups**
:::

------------------------------------------------------------------------

### One-Way ANOVA : Definition

<br> <br>

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
<p align="center">

<strong>One-Way ANOVA</strong>

</p>

-   one-way refers to the number of explanatory factors in the ANOVA model

one-way ANOVA has:

-   one continuous outcome
-   one categorical factor
-   two or more factor levels
:::

------------------------------------------------------------------------

### One-Way ANOVA Model

::: {style="font-size:0.72em"}
One-way ANOVA model:

$$
y_{ij} = \mu + \alpha_i + \epsilon_{ij}
$$

where:

-   $y_{ij}$ : outcome for observational unit $j$ in group $i$
-   $\mu$ : overall mean level
-   $\alpha_i$ : effect of group $i$
-   $\epsilon_{ij}$ : residual variation for observational unit $j$ in group $i$
:::

------------------------------------------------------------------------

### One-Way ANOVA : Hypotheses

<br>

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
**Null hypothesis (H₀)**: all population group means are equal.

$$
\mu_1 = \mu_2 = \mu_3 = \cdots = \mu_k
$$

**Alternative hypothesis (**$H_A$): at least one population group mean differs from another
:::

<br>

::: {style="font-size:0.72em"}
Note: alternative hypothesis does not state which groups differ, instead, states that the group means are not all equal
:::

------------------------------------------------------------------------

### Key Definition : Outcome Variable

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
<p align="center">

<strong>Outcome Variable</strong>

</p>

**Outcome variable** is the measured quantity being compared across groups
:::

<br>

::: {style="font-size:0.74em"}
In one-way ANOVA, the outcome is usually continuous

Biomedical examples include:

-   systolic blood pressure
-   PSA level
-   cholesterol level
-   tumor volume
-   gene expression value
-   biomarker concentration
:::

------------------------------------------------------------------------

### Key Definition : Factor

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
<p align="center">

<strong>Factor</strong>

</p>

A **factor** is a categorical explanatory variable that assigns each observational unit to a group.
:::

<br>

::: {style="font-size:0.74em"}
Examples of factors include:

-   treatment group
-   disease subtype
-   tumor stage
-   genotype category
-   hospital unit
-   exposure category

In one-way ANOVA, there is one factor
:::

------------------------------------------------------------------------

### Key Definition : Levels

::: {style="background-color:#e6f3ff; padding:1em; border-radius:6px; border:1px solid #4a90e2;"}
<p align="center">

<strong>Levels</strong>

</p>

The **levels** of a factor are the specific groups being compared.
:::

<br>

::: {style="font-size:0.74em"}
If the factor is treatment group, the levels might be:

-   placebo
-   low-dose drug
-   high-dose drug

One-way ANOVA evaluates whether the population means for these levels appear equal based on the observed data
:::

------------------------------------------------------------------------

### One-Way ANOVA: Example Scientific Questions

::: {style="font-size:0.74em"}
One-way ANOVA can be used when the scientific question asks whether a continuous measurement differs across categories of one grouping variable.

Examples:

-   Do mean PSA levels differ across tumor stage groups?
-   Do mean biomarker concentrations differ across treatment groups?
-   Do mean tumor volumes differ across genotype categories?
-   Do mean blood pressure values differ across lifestyle intervention groups?

Note: each question has one continuous outcome and one categorical factor
:::

------------------------------------------------------------------------

### Comparison: One-Way ANOVA and t-Test

::: {style="font-size:0.62em"}
| Feature | Independent-samples t-test | One-way ANOVA |
|:---|:---|:---|
| **Outcome** | Continuous | Continuous |
| **Number of groups** | Two groups | Two or more groups |
| **Main test statistic** | t statistic | F statistic |
| **Null hypothesis** | Two population means are equal | All population group means are equal |
| **Primary result** | Evidence about one mean difference | Evidence that at least one group mean differs |
:::

------------------------------------------------------------------------

### Why Not Run Many t-Tests?

::: {style="font-size:0.74em"}
Suppose a study compares mean biomarker levels across four treatment groups

Pairwise t-tests would compare:

-   group 1 vs group 2
-   group 1 vs group 3
-   group 1 vs group 4
-   group 2 vs group 3
-   group 2 vs group 4
-   group 3 vs group 4

Possible Outcome:

-   each test creates another opportunity for a false positive result

-   when there are exactly two groups: one-way ANOVA and an independent-samples t-test evaluate the same mean comparison

-   ANOVA extends the mean-comparison logic of the t-test to more than two groups
:::

------------------------------------------------------------------------

### Biomedical Example Dataset

::: {style="font-size:0.74em"}
Scientific Question:

<br>

<p align="center">

> <strong>Do mean inflammatory biomarker levels differ across treatment groups?</strong>

</p>

<br>

Note:

-   outcome : biomarker level

-   factor : treatment group

-   levels : placebo, standard dose, and high dose.
:::

------------------------------------------------------------------------

### Simulate Example Data in R

```{r, echo=FALSE}
anova_data <- tibble(
  treatment = rep(c("Placebo", "Standard dose", "High dose"), each = 36),
  biomarker = c(
    rnorm(36, mean = 52, sd = 8),
    rnorm(36, mean = 47, sd = 8),
    rnorm(36, mean = 42, sd = 8)
  )
)

anova_data <- anova_data |>
  mutate(treatment = factor(treatment, levels = c("Placebo", "Standard dose", "High dose")))

glimpse(anova_data)
```

::: {style="font-size:0.66em"}
-   three groups with 36 observations per group
-   `treatment` variable : coded as a factor because it defines group membership
-   `biomarker` variable : numeric because it is the continuous outcome being compared across groups
:::

------------------------------------------------------------------------

### Inspect Group Sizes

```{r, echo=FALSE}
anova_data |>
  count(treatment)
```

::: {style="font-size:0.68em"}
-   check the number of observations in each group

-   group size affects the precision of group mean estimates
:::

------------------------------------------------------------------------

### Visualize Data Before Testing

```{r, fig.width=8, fig.height=5, echo=FALSE}
ggplot(anova_data, aes(x = treatment, y = biomarker)) +
  geom_boxplot(width = 0.45, outlier.shape = NA) +
  geom_jitter(width = 0.10, alpha = 0.60) +
  labs(
    title = "Biomarker levels by treatment group",
    subtitle = "One-way ANOVA compares the group means relative to within-group variability",
    x = "Treatment group",
    y = "Biomarker level"
  )
```

::: {style="font-size:0.66em"}
-   plot shows both the center and spread of the biomarker values within each treatment group
-   ANOVA uses this information: how far group means are from each other and how much values vary within each group to calculate the F-statistic, a signal-to-noise ratio that determines whether the differences between group means are statistically significant
:::

------------------------------------------------------------------------

### Estimate Group Means

```{r group-means}
group_summary <- anova_data |>
  group_by(treatment) |>
  summarize(
    n = n(),
    mean_biomarker = mean(biomarker),
    sd_biomarker = sd(biomarker),
    .groups = "drop"
  )

group_summary
```

::: {style="font-size:0.66em"}
-   group means provide the descriptive comparison
-   one-way ANOVA evaluates whether differences among these sample means are large enough relative to within-group variation to provide evidence against equal population means
:::

------------------------------------------------------------------------

### Variation Within the Model

::: {style="font-size:0.74em"}
**one-way ANOVA** : asks whether group membership explains enough outcome variation to justify rejecting the equal-means null hypothesis

The total variation is separated into:

-   **between-group variation**: how much the group means differ from the overall mean
-   **within-group variation**: how much individual observations differ from their own group mean

**F statistic** : compares these two sources of variation after accounting for their degrees of freedom
:::

------------------------------------------------------------------------

### Manual Pieces of the ANOVA Calculation

```{r manual-ss}
overall_mean <- mean(anova_data$biomarker)

ss_between <- anova_data |>
  group_by(treatment) |>
  summarize(n = n(), group_mean = mean(biomarker), .groups = "drop") |>
  summarize(ss_between = sum(n * (group_mean - overall_mean)^2)) |>
  pull(ss_between)

ss_within <- anova_data |>
  group_by(treatment) |>
  summarize(ss_within = sum((biomarker - mean(biomarker))^2), .groups = "drop") |>
  summarize(ss_within = sum(ss_within)) |>
  pull(ss_within)

c(ss_between = ss_between, ss_within = ss_within)
```

::: {style="font-size:0.62em"}
-   code calculates two sources of variation
-   `ss_between` : measures how far group means are from the overall mean
-   `ss_within` : measures how far individual observations are from their own group mean
:::

------------------------------------------------------------------------

### Manual F Statistic

```{r manual-f}
k <- nlevels(anova_data$treatment)
n_total <- nrow(anova_data)

df_between <- k - 1
df_within <- n_total - k

ms_between <- ss_between / df_between
ms_within <- ss_within / df_within

f_statistic <- ms_between / ms_within
p_value <- pf(f_statistic, df_between, df_within, lower.tail = FALSE)

c(
  df_between = df_between,
  df_within = df_within,
  ms_between = ms_between,
  ms_within = ms_within,
  F = f_statistic,
  p_value = p_value
)
```

::: {style="font-size:0.62em"}
-   F statistic : between-group mean square divided by the within-group mean square

$$
F = \frac{\text{variance between groups}}{\text{variance within groups}}
$$

-   **larger F statistic** : indicates that differences among group means are large relative to the variability observed inside the groups
:::

------------------------------------------------------------------------

### Fit One-Way ANOVA in R

```{r }
anova_fit <- aov(biomarker ~ treatment, data = anova_data)

summary(anova_fit)
```

::: {style="font-size:0.66em"}
-   `biomarker ~ treatment` : biomarker level is modeled as a function of treatment group
-   `treatment` : factor $\rightarrow$ R fits a model that estimates differences among treatment-group means
:::

------------------------------------------------------------------------

### ANOVA Table

::: {style="font-size:0.68em"}
ANOVA table usually includes:

-   **Df**: degrees of freedom associated with the factor and residual variation
-   **Sum Sq**: variation attributed to the factor and residual error
-   **Mean Sq**: sum of squares divided by degrees of freedom
-   **F value**: ratio of factor mean square to residual mean square
-   **Pr(\>F)**: p-value for the overall ANOVA F-test

The treatment p-value evaluates the null hypothesis that all treatment-group population means are equal
:::

------------------------------------------------------------------------

### ANOVA Table

::: {style="font-size:0.68em"}
ANOVA table usually includes:

-   **Df**: degrees of freedom associated with the factor and residual variation
-   **Sum Sq**: variation attributed to the factor and residual error
-   **Mean Sq**: sum of squares divided by degrees of freedom
-   [**F value**: ratio of factor mean square to residual mean square]{style="color: #0076B6;"}
-   [**Pr(\>F)**: p-value for the overall ANOVA F-test]{style="color: #0076B6;"}
:::

------------------------------------------------------------------------

### Fit One-Way ANOVA in R

```{r }
anova_fit <- aov(biomarker ~ treatment, data = anova_data)

summary(anova_fit)
```

::: {style="font-size:0.66em"}
-   ANOVA tells you at least one group mean differs

-   model indicates that treatment has a statistically significant effect on biomarker levels:

    -   F(2, 105) = 13.17
        -   estimated variability due to treatment differences is about 13 times larger than the average unexplained variability within treatment groups
    -   p = 7.88 × 10⁻⁶

Note: to identify the specific group differences, perform a post-hoc multi-comparison test \[Tukey's Honest Significant Difference (HSD) test\]
:::

------------------------------------------------------------------------

### ANOVA: Check Model Assumptions

1.  **normality**: observations in group i are assumed to come from a normal distribution
2.  **equal (constant) variance**: each group is assumed to have the same population variance
3.  **independence**: observations should be independent of one another

------------------------------------------------------------------------

### ANOVA Model Assumptions: Residuals by Treatment Group

::: {style="font-size:0.66em"}
```{r, fig.width=8, fig.height=5, echo=FALSE}
boxplot(
  residuals(anova_fit) ~ anova_data$treatment,
  xlab = "Treatment",
  ylab = "Residuals",
  main = "Residuals by Treatment Group"
)

abline(h = 0, lty = 2)
```

-   not a formal diagnostic plot

-   Goal: visually assess whether group variances are reasonably similar

-   **residual** : error or difference between an actual observed data point and the value predicted by a statistical model

-   residuals centered near zero in each group

-   similar box heights (similar variability)

-   no group with substantially larger spread
:::

------------------------------------------------------------------------

### ANOVA Model Assumptions: Equal Variance \[Homoskedacity\]

::: {style="font-size:0.66em"}
```{r, fig.width=8, fig.height=5, echo=FALSE}
plot(
  anova_fit,
  which = 1
)
```

-   **residual** : error or difference between an actual observed data point and the value predicted by a statistical model

residuals vs Fitted (which = 1) — introduces residual diagnostics

-   residuals centered around zero
-   no obvious curve
-   similar spread across fitted values
:::

------------------------------------------------------------------------

### ANOVA Model Assumptions: Test for Normality

```{r, fig.width=8, fig.height=5, echo=FALSE}
plot(
  anova_fit,
  which = 2
)
```

::: {style="font-size:0.66em"}
normal Q-Q Plot (which = 2) — assesses the normality assumption

-   points lying close to the reference line
-   only small deviations at the tails
:::

------------------------------------------------------------------------

### ANOVA Model Assumptions: Independence

::: {style="font-size:0.66em"}
```{r, fig.width=8, fig.height=5, echo=FALSE}
# extract the residuals
res <- residuals(anova_fit)
#plot
plot(
  seq_along(res),
  res,
  pch = 19,
  xlab = "Observation Order",
  ylab = "Residuals",
  main = "Residuals vs Observation Order"
)

abline(h = 0, lty = 2, col = "red")
```

Test: plot maps model errors against the chronological sequence or physical order in which your data was collected

scale-Location Plot (which = 3) — assesses homoscedasticity

Evidence supporting independence:

-   points randomly scattered around zero
-   no upward or downward trend
:::

------------------------------------------------------------------------

### Summary

-   introduced one-way ANOVA as a method for comparing means of a continuous outcome across groups
-   one-way ANOVA extends the two-sample t-test to studies involving two or more groups
-   F-statistic : constructed by comparing between-group and within-group variation
-   demonstrated how to fit a one-way ANOVA model and interpret the model output

------------------------------------------------------------------------
